Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-10 ms.tex
sigma_bullet

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to
measure the radial surface brightness profiles for hundreds of nearby
early type galaxies. The measured profile is fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law
slope, $\Gamma$, to an outer power law slope, $\beta$, at a
break radius, $\rb$.  If the stellar population is spherically
symmetric, then this corresponds to a 3D stellar density $\rhostar \propto
r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for
$r \gg \rb$.  We can write the deprojected stellar density approximately (formally, this is the $\alpha \rightarrow \infty$ limit) as
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rinf} \left(r/\rinf\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rinf}$ is the stellar density at the radius of the black hole sphere of influence\footnote{This approximate expression agrees surprisingly well with the mean empirical scaling relation for $r_{\rm inf}(M_\bullet)$ calculated in Appendix C of \citet{Stone&Metzger15}, although we note that this relation has significant scatter.}, 
\begin{equation}
\rinf \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\end{equation}
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the
second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$
relationship of \citet{McConnellMa+:2011a},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of an obviously better alternative.

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}), where
\be\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2},
\label{eq:sigmarel}
\ee 
is the stellar velocity dispersion and which accounts for the minimal amount of shock heating from stellar winds due to the random motion of stars in the SMBH potential, while $\sigma_0^{2} \approx 3 \sigma^2$ is assumed to be constant, corresponding to the observed
velocity dispersion outside of the SMBH sphere of influence
\citep{ChaeBernardi+:2014a}.  The second term, $v_{w}^{2}$, parameterizes additional sources of energy input, including faster winds from young
stars, millisecond pulsars, supernovae, AGN feedback, etc  ($\S\ref{sec:heating}$).  We assume
that $v_w$ is constant with radius, i.e. that the volumetric heating
rate is proportional to the local stellar density.  
%The value of $v_w$
%will in general depend on the SMBH mass and the age of the stellar
%population ($\S\ref{sec:heating}$).

@@

The accretion rate onto the SMBH is given by the total mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& 4\pi \int_{0}^{r_{s}}q r^{2}dr = \frac{\eta M_{\star}|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    4.5 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    3.2 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have assumed $v_{w} \gg \sigma$ by adopting equation
(\ref{eq:stag_simple}) for $r_s$.  The resulting Eddington ratio is
given by
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    2.0 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    1.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$ is the
Eddington accretion rate, assuming a radiative efficiency of ten per
cent.  Note the sensitive dependence of the accretion rate on the wind
heating rate.
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

@@

\begin{equation}
\frac{r_{\rm circ}}{r_{s}} \approx \frac{r_{s}}{r_{\rm inf}}\lambda_{R}^{2} \lesssim \lambda_{R}^{2},
\label{eq:rcirc}
\end{equation}
where we have used the definition $r_{\rm inf} \equiv
GM_{\bullet}/\sigma^{2}$ and in the second inequality have assumed
that $r_{\rm s} \lesssim r_{\rm inf}$, a condition which is satisfied
for the thermally-stable solutions of interest.

@@

  & & & Cusp Galaxies, $\Gamma^{(a)} = 0.8$ & & & & & \\

@@

& & & Core Galaxies, $\Gamma = 0.1$ & & & & & & \\

@@

The high value of $v_{w}^{\rm Ia}$ implies that Ia SNe represent an
important source of CNM heating.  However, SNe can only be
approximated as supplying heating which is spatially and temporally
homogeneous if the rate of SNe is rapid compared to the characteristic
evolution time of the flow at the radius of interest
(\citealt{ShcherbakovWong+:2014a}).  We define the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma_0}\right)^{1/2} \sim
    29 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
  as the location exterior to which the time interval between
  subsequent supernovae $\tau_{\rm Ia} \sim (M_{\rm enc}R_{\rm
    Ia})^{-1} \sim G/(r\sigma_0^{2}R_{\rm Ia})$ exceeds the local
  dynamical timescale $t_{\rm dyn} \sim r/\sigma_0$, where we again
  adopt the Ia rate for an old stellar population and in the final
  equality we estimate $\sigma_0 \approx \sqrt{3}\sigma$ using $M_{\bullet}-\sigma$ (eq.~[\ref{eq:Msigma}]).

>>>>

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to
measure the radial surface brightness profiles for hundreds of nearby
early type galaxies. The measured profile is fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law
slope, $\Gamma$, to an outer power law slope, $\beta$, at a
break radius, $\rb$.  If the stellar population is spherically
symmetric, then this corresponds to a 3D stellar density $\rhostar \propto
r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for
$r \gg \rb$.  We can write the deprojected stellar density approximately (formally, this is the $\alpha \rightarrow \infty$ limit) as
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rinf} \left(r/\rinf\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rinf}$ is the stellar density at the radius of the black hole sphere of influence\footnote{This approximate expression agrees surprisingly well with the mean empirical scaling relation for $r_{\rm inf}(M_\bullet)$ calculated in Appendix C of \citet{Stone&Metzger15}, although we note that this relation has significant scatter.}, 
\begin{equation}
  \rinf \simeq G \Mbh/\sigma_{\bullet}^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\end{equation}
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the
second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$
relationship of \citet{McConnellMa+:2011a},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma_{\bullet}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g., \citealt{Greene&Ho07}), we adopt equation (\ref{eq:Msigma}) for lack of an obviously better alternative.

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}), where

\begin{equation}
\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2},
\label{eq:sigmarel}
\end{equation}

is the stellar velocity dispersion and which accounts for the minimal
amount of shock heating from stellar winds due to the random motion of
stars in the SMBH potential. We take $\sigma_0$ to be constant and use
$\sigma_0^{2} \approx 3 \sigma_{\bullet}^2$, where
$\sigma_{\bullet}\simeq 170 \Mbheight^{0.2}$ km/s is the velocity
dispersion from the \citet{McConnellMa+:2011a} $\Mbh-\sigma$
relation. 
% \citep{ChaeBernardi+:2014a}
.  The second term, $v_{w}^{2}$, parameterizes additional sources of
energy input, including faster winds from young stars, millisecond
pulsars, supernovae, AGN feedback, etc ($\S\ref{sec:heating}$).  We
assume that $v_w$ is constant with radius, i.e. that the volumetric
heating rate is proportional to the local stellar density.
%The value of $v_w$
%will in general depend on the SMBH mass and the age of the stellar
%population ($\S\ref{sec:heating}$).

@@

The accretion rate onto the SMBH is given by the total mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& 4\pi \int_{0}^{r_{s}}q r^{2}dr = \frac{\eta M_{\star}|_{r_{\rm s}}}{\th} \nonumber \\
&\approx&
  \begin{cases}
    4.5 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    3.2 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have assumed $v_{w} \gg \sigma_0$ by adopting equation
(\ref{eq:stag_simple}) for $r_s$.  The resulting Eddington ratio is
given by
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    2.0 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core}, \\
    1.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$ is the
Eddington accretion rate, assuming a radiative efficiency of ten per
cent.  Note the sensitive dependence of the accretion rate on the wind
heating rate.
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

@@

\begin{equation}
\frac{r_{\rm circ}}{r_{s}} \approx \frac{r_{s}}{r_{\rm inf}}\lambda_{R}^{2} \lesssim \lambda_{R}^{2},
\label{eq:rcirc}
\end{equation}
where we have used the definition $r_{\rm inf} \equiv
GM_{\bullet}/\sigma_{\bullet}^{2}$ and in the second inequality have assumed
that $r_{\rm s} \lesssim r_{\rm inf}$, a condition which is satisfied
for the thermally-stable solutions of interest.

@@

 Cusp Galaxies, $\Gamma^{(a)} = 0.8$ & & & & & & & & \\

@@

Core Galaxies, $\Gamma = 0.1$  &  & & & & & & & & \\

@@

The high value of $v_{w}^{\rm Ia}$ implies that Ia SNe represent an
important source of CNM heating.  However, SNe can only be
approximated as supplying heating which is spatially and temporally
homogeneous if the rate of SNe is rapid compared to the characteristic
evolution time of the flow at the radius of interest
(\citealt{ShcherbakovWong+:2014a}).  We define the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma_0}\right)^{1/2} \sim
    29 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
  as the location exterior to which the time interval between
  subsequent supernovae $\tau_{\rm Ia} \sim (M_{\rm enc}R_{\rm
    Ia})^{-1} \sim G/(r\sigma_0^{2}R_{\rm Ia})$ exceeds the local
  dynamical timescale $t_{\rm dyn} \sim r/\sigma_0$, where we again
  adopt the Ia rate for an old stellar population and in the final
  equality we estimate $\sigma_0 \approx \sqrt{3}\sigma_{\bullet}$ using $M_{\bullet}-\sigma$ (eq.~[\ref{eq:Msigma}]).
